Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-10-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:02	17:09:02	(Fri 2.Oct)

\author{Jo~Inge~Buskenes, %
        Roy~Edgar~Hansen, %
        Andreas~Austeng%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department of Informatics, University of Oslo, Norway.}%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.E.Hansen is also with The Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space

@@

This article is outlined as follows: In section \ref{beamforming} we introduce beamforming, \ref{mvdr} MVDR \ref{lca} LCA \ref{} \ref{} \ref{}

@@

\newpage
\subsection{Receive Beamforming}\label{beamforming}

@@

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window or a taper function. Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}. Real windows lead to symmetric responses, while complex weights allow asymmetric responses.

@@

\section{MVDR}\label{mvdr}

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}. To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\section{LCA}\label{lca}

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
%
\begin{align}
\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights. As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

@@

\subsection{Window function: Kaiser}\label{kaiser_windows}

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

\subsection{Steering}

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{steering_vector}
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

\subsection{Steering bounds}\label{steering_bounds}

@@

We call the upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{steering_kaiser_3dB}
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

\subsection{Trigonometric windows}

While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
%
\begin{align}
f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{trig_window_function}
\end{align}
%
where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients. Steering is applied as in (\ref{steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{steering_bounds}:
%
\begin{align}
\phi_\text{3dB}(\alpha) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136.
\end{align}

\subsection{Oversampling}

@@

\section{Results and Discussion}\label{results_discussion}

@@

The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows. In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}). In section \ref{lca_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).

\subsection{LCA window function}\label{lca_window_function}

@@

\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{lca_kaiser_parameters}

@@

\subsection{Window database size}\label{lca_database_size}

@@

\subsection{Image quality}

@@

\subsection{Computational complexity}

@@

In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs. Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access. If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{reducing_lca_complexity}. This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.

@@

\section{Conclusion}

@@

\subsection{Summary}

@@

\section{Reducing LCA complexity}\label{reducing_lca_complexity}

@@

Throughout this article we have used LCA with the Kaiser window function. However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function. For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. To derive it, we start by inserting the trigonometric function (\ref{trig_window_function}), with applied steering (\ref{steering_vector}), into the beamformer equation (\ref{z}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
\end{align}
%
where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{steering_vector}), and 
%
\begin{align}
\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
Unity gain in the look direction is ensured as long as the weights sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
\end{align}
%
This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.

>>>>

\author{Jo~Inge~Buskenes, %
        Roy~Edgar~Hansen, %
        Andreas~Austeng%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department of Informatics, University of Oslo, Norway.}%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.~E.~Hansen is also with the Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space

@@

This article is outlined as follows: In Section \ref{sec:beamforming} we introduce beamforming, followed by background on the MVDR beamformer in Section \ref{sec:mvdr}. In \ref{sec:lca} we provide complete background on the LCA method, by presenting the Kaiser window function in \ref{sec:lca_kaiser_windows}, the method for steering it in \ref{sec:lca_steering} as well as a note on maximum steering amount in \ref{sec:lca_steering_bounds}, and a mention of the alternative trigonometric window function in \ref{sec:lca_trigonometric} and sampling considerations in \ref{sec:lca_oversampling}.

@@

\newpage
\section{Receive Beamforming}\label{sec:beamforming}

@@

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window or a taper function. Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}. Real windows lead to symmetric responses, while complex weights allow asymmetric responses.

@@

\section{MVDR}\label{sec:mvdr}

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}. To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\section{LCA}\label{sec:lca}

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
%
\begin{align}
\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights. As will be demonstrated in section \ref{sec:results_lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

@@

\subsection{Window function: Kaiser}\label{sec:lca_kaiser_windows}

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{eq:kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

\subsection{Steering}\label{sec:lca_steering}

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{eq:steering_vector}
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

\subsection{Steering bounds}\label{sec:lca_steering_bounds}

@@

We call the upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

\subsection{Trigonometric windows}\label{sec:lca_trigonometric}

While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
%
\begin{align}
f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
\end{align}
%
where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients. Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{sec:lca_steering_bounds}:
%
\begin{align}
\phi_\text{3dB}(\alpha) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136.
\end{align}

\subsection{Oversampling}\label{sec:lca_oversampling}

@@

\section{Results and Discussion}\label{sec:results_discussion}

@@

The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows. In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}). In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).

\subsection{LCA window function}\label{sec:results_lca_window_function}

@@

\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}

@@

\subsection{Window database size}\label{sec:results_database_size}

@@

\subsection{Image quality}\label{sec:results_image_quality}

@@

\subsection{Computational complexity}\label{sec:results_complexity}

@@

In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs. Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access. If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}. This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.

@@

\section{Conclusion}\label{sec:conclusion}

@@

\subsection{Summary}\label{sec:summary}

@@

\section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}

@@

Throughout this article we have used LCA with the Kaiser window function. However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function. For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. To derive it, we start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
\end{align}
%
where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
%
\begin{align}
\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
Unity gain in the look direction is ensured as long as the weights sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
\end{align}
%
This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
